
For this example, we will create a fake dataset that contains the following two variables for 15 students:

Total hours studied for some exam
Exam score
We will attempt to fit a simple linear regression model using hours as the explanatory variable and exam score as the response variable.



### Create dataset


```{r}


df <- data.frame(hours=c(1, 2, 4, 5, 5, 6, 6, 7, 8, 10, 11, 11, 12, 12, 14),
                 score=c(64, 66, 76, 73, 74, 81, 83, 82, 80, 88, 84, 82, 91, 93, 89));

head(df)

attach(df)

scatter.smooth(hours, score, main ='Hours studied vs. Exam Score')

```

```{r}
boxplot(score)
```

Next, we can create a boxplot to visualize the distribution of exam scores and check for outliers. By default, R defines an observation to be an outlier if it is 1.5 times the interquartile range greater than the third quartile (Q3) or 1.5 times the interquartile range less than the first quartile (Q1).

If an observation is an outlier, a tiny circle will appear in the boxplot:

### Fit simple linear regression model

```{r}
model <- lm(score~hours)

```

### View model summary

```{r}
summary(model)
```

From the model summary we can see that the fitted regression equation is:

Score = 65.334 + 1.982*(hours)

This means that each additional hour studied is associated with an average increase in exam score of 1.982 points. And the intercept value of 65.334 tells us the average expected exam score for a student who studies zero hours.

We can also use this equation to find the expected exam score based on the number of hours that a student studies. For example, a student who studies for 10 hours is expected to receive an exam score of 85.15

Score = 65.334 + 1.982*(10) = 85.15

Here is how to interpret the rest of the model summary:

Pr(>|t|): This is the p-value associated with the model coefficients. Since the p-value for hours (2.25e-06) is significantly less than .05, we can say that there is a statistically significant association between hours and score.
Multiple R-squared: This number tells us the percentage of the variation in the exam scores can be explained by the number of hours studied. In general, the larger the R-squared value of a regression model the better the explanatory variables are able to predict the value of the response variable. In this case, 83.1% of the variation in scores can be explained hours studied.
Residual standard error: This is the average distance that the observed values fall from the regression line. The lower this value, the more closely a regression line is able to match the observed data. In this case, the average observed exam score falls 3.641 points away from the score predicted by the regression line.
F-statistic & p-value: The F-statistic (63.91) and the corresponding p-value (2.253e-06) tell us the overall significance of the regression model, i.e. whether explanatory variables in the model are useful for explaining the variation in the response variable. Since the p-value in this example is less than .05, our model is statistically significant and hours is deemed to be useful for explaining the variation in score.

### Create Residual Plots
After we have fit the simple linear regression model to the data, the last step is to create residual plots.

One of the key assumptions of linear regression is that the residuals of a regression model are roughly normally distributed and are homoscedastic at each level of the explanatory variable. If these assumptions are violated, then the results of our regression model could be misleading or unreliable.

To verify that these assumptions are met, we can create the following residual plots:

Residual vs. fitted values plot: This plot is useful for confirming homoscedasticity. The x-axis displays the fitted values and the y-axis displays the residuals. As long as the residuals appear to be randomly and evenly distributed throughout the chart around the value zero, we can assume that homoscedasticity is not violated

```{r}
#define residuals
res <- resid(model)

#produce residual vs. fitted plot
plot(fitted(model), res)

#add a horizontal line at 0 
abline(0,0)
```
The residuals appear to be randomly scatted around zero and don’t exhibit any noticeable patterns, so this assumption is met.

Q-Q plot: This plot is useful for determining if the residuals follow a normal distribution. If the data values in the plot fall along a roughly straight line at a 45-degree angle, then the data is normally distributed

```{r}
#create Q-Q plot for residuals
qqnorm(res)

#add a straight diagonal line to the plot
qqline(res) 
```


The residuals stray from the 45-degree line a bit, but not enough to cause serious concern. We can assume that the normality assumption is met.

Since the residuals are normally distributed and homoscedastic, we have verified that the assumptions of the simple linear regression model are met. Thus, the output from our model is reliable.